---
title: "p8105_hw3_ryw2109"
author: "Rita Wang"
date: "2024-10-16"
output: github_document
---

```{r default_setting, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(ggthemes)
library(ggridges)
library(patchwork)
# Importing data
  # devtools::install_github("p8105/p8105.datasets") --> installing dataset for P8105
library(p8105.datasets)


knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```

### Question 1
```{r Q1_dataset_A}
data("ny_noaa") # calling for ny_noaa data
# ny_noaa #displays ny_noaa

nynoaa_df = ny_noaa %>% 
  separate(date, into = c("year", "month", "day"), convert = TRUE) %>% 
    # Create separate variables for year, month, and day
  mutate( # Ensure observations for temperature, precipitation, and snowfall are given in reasonable units
    tmax = as.numeric(tmax),
    tmin = as.numeric(tmin)
    )

nynoaa_df %>% 
  count(snow) %>%
  arrange(desc(n))
```
There is a total of `r nrow(nynoaa_df)` rows and `r ncol(nynoaa_df)` columns. The following are the variable names of the dataset:

`r names(nynoaa_df)`

Reasonable units for maximum and minimum temperature is Celsius, and reasonable units for precipitation, snowfall, and snowfall depth is mm. For snowfall, the most commonly observed value is 0 mm to which there were 2008508 observed data values; this is because snow does not commonly occur throughout the year.

```{r Q1_B}
# Make a two-panel plot showing the average max temperature in January and in July in each station across years
nynoaa_df %>% 
  group_by(
    id, year, month
    ) %>% 
  filter(
    month %in% c(1, 7) #January and in July
    ) %>% 
  summarize(
    mean_tmax = mean(tmax, na.rm = TRUE) # average max temperature
    ) %>% 
  ggplot(aes(x = year, y = mean_tmax, group = id, color = month)) + 
  geom_point(aes(alpha = 0.3)) + 
  facet_grid(. ~ month) + #two-panel plot 
  labs(
    title = "Mean monthly temperature for each station across years for January and July",
    x = "Month",
    y = "Temperature (Celsius)"
  )
```

The average maximum temperature in January is much lower than that of July. There is one outlier in July around 1985 to which the temperature is slightly colder than the other average maximum temperature in July.

```{r Q1_C}
hex_nynoaa_df = nynoaa_df %>% 
  ggplot(aes (x = tmin , y = tmax) ) + # tmax vs tmin for the full dataset
  geom_hex()

ridges_nynoaa_df = nynoaa_df %>% 
  filter(snow < 100, snow > 0) %>% # snowfall values greater than 0 and less than 100
  ggplot(aes (x = snow , y = as.factor(year))) + # separately by year.
  geom_density_ridges() # distribution

hex_nynoaa_df + ridges_nynoaa_df #two-panel plot
```

### Question 2
```{r Q2_A}
demograph = read_csv(file = "./data/nhanes_covar.csv", skip = 4) %>%  #import data
    # rows 1-3 had legends
  janitor::clean_names()  #clean column name

demograph

accelerometer = read_csv(file = "./data/nhanes_accel.csv") %>%  #import data
  janitor::clean_names()  #clean column name

accelerometer

demo_accel = left_join(demograph , accelerometer , by = "seqn") %>% 
    mutate ( # encode data with reasonable variable classes
    seqn = as.integer(seqn),
    sex = as.integer(sex),
    age = as.integer(age),
    education = as.integer(education)
  ) %>%
  filter(age >= 21) %>% # excluding participants less than 21 years of age
  drop_na("sex", "age", "bmi") # excluding those with missing demographic data

demo_accel

```

After merging the two datasets, there is now a total of `r nrow(demo_accel)` rows and `r ncol(demo_accel)` columns. 

It can be seen that the original data set had seqn, sex, age, and education recorded as dbls, however, they are integers.

```{r Q2_B}
# table for the number of men and women in each education category
demo_accel %>% 
  group_by(sex, education) %>% 
  summarize(total_count = n()) %>% 
  pivot_wider(
    names_from = "education",
    values_from = total_count
  ) %>% 
  knitr::kable(digits = 1)

# age distributions for men and women in each education category
demo_accel %>%
  ggplot(aes(x = age, fill = as.factor(sex))) +
  geom_histogram(position = "dodge") +
  facet_grid(. ~ education) +
  labs(title = "Age Distribution For Sex VS Education",
       x = "Age",
       y = "Count"
  )

```

Based on the histogram, among those with an education less than high school, it can be seen there is a larger number of females and males who are of an older age (60 and above). Based on the table created, there are 27 males and 28 females who have an education less than high school. 

Based on the histogram, among those with an education equivalent to high school, it can be seen there is a relatively even spread across the different ages. Based on the table created, there are 35 males and 23 females who have an education if high school equivalent. 

Based on the histogram, among those with an education more than high school, it can be seen there is a larger number of females and males who are of a younger age (40 and below). Based on the table created, there are 56 males and 59 females who have an education more than high school.

Across the three histograms, it can be seen that there are more individuals who have an education more than high school.

```{r Q2_C}
total_demo_accel = demo_accel %>%
  mutate(
    total_activities = rowSums(select(., starts_with("min"))) # sums all the values of min1:min1440 for each participant
  ) %>% 
  select(
    seqn, sex, age, bmi, education, total_activities
    ) 

total_demo_accel %>% 
  ggplot(aes(x = age, y = total_activities, color = sex)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ education) +
  labs(title = "Total Activity By Age For Sex VS Education",
       x = "Age",
       y = "Total Activity"
  )
```

From the graphs, we can see that for all 3 educational levels, there is a general decrease of activity as age increases.

For individuals with an education less than high school, older males had higher total activity compared to females. Between age 50 to 60, the originally decreasing total activity with age had an increase.

For individuals with an education equivalent to high school, younger and older females had higher total activity compared to males. Between age 30 to 40, there seems to be an increase of total activity. At age 40, the total activity decreased and leveled off starting at age 55.

For individuals with an education higher than high school, older females had higher total activity compared to males. In general, there is a general decrease in total activity as age increases.

```{r Q2_D}
long_hour_demo_accel = demo_accel %>%
  select(seqn, sex, age, bmi, education, min1:min1440) %>% 
  mutate( # converting minutes to hours
    hour1 = rowSums(select(., min1:min60)),
    hour2 = rowSums(select(., min61:min120)),
    hour3 = rowSums(select(., min121:min180)),
    hour4 = rowSums(select(., min181:min240)),
    hour5 = rowSums(select(., min241:min300)),
    hour6 = rowSums(select(., min301:min360)),
    hour7 = rowSums(select(., min361:min420)),
    hour8 = rowSums(select(., min421:min480)),
    hour9 = rowSums(select(., min481:min540)),
    hour10 = rowSums(select(., min541:min600)),
    hour11 = rowSums(select(., min601:min660)),
    hour12 = rowSums(select(., min661:min720)),
    hour13 = rowSums(select(., min721:min780)),
    hour14 = rowSums(select(., min781:min840)),
    hour15 = rowSums(select(., min841:min900)),
    hour16 = rowSums(select(., min901:min960)),
    hour17 = rowSums(select(., min961:min1020)),
    hour18 = rowSums(select(., min1021:min1080)),
    hour19 = rowSums(select(., min1081:min1140)),
    hour20 = rowSums(select(., min1141:min1200)),
    hour21 = rowSums(select(., min1201:min1260)),
    hour22 = rowSums(select(., min1261:min1320)),
    hour23 = rowSums(select(., min1321:min1380)),
    hour24 = rowSums(select(., min1381:min1440))
  )

hour_demo_accel = long_hour_demo_accel %>% 
  select(seqn, sex, age, bmi, education, hour1:hour24) %>% 
  pivot_longer( # converts the hour1:hour24 columns into rows
    hour1:hour24,
    names_to = "hr",
    names_prefix = "",
    values_to = "hour_activity"
    ) %>% 
  mutate(
    hr = gsub("hour", "", hr),
    hr = as.integer(hr)
  )

hour_demo_accel %>% 
  ggplot(aes(x = hr, y = hour_activity, color = sex)) +
  geom_point(aes(alpha = 0.5)) +
  facet_grid(. ~ education) +
  labs(title = "Hourly Activity For Sex VS Education",
       x = "Hours",
       y = "Activity"
  )
```

The average activity of a 24 hour time period have a relatively similar spread of across all 3 education levels, with a low amount of activity within in the first 5 hours of a day to which the total activity increases peaking at hour 10-15. Males and females with an education less than high school have a relatively similar spread of exercise. For those with an education equivalent to high school, we see that the data points are darker for each hour, indicating that more males are having activities than females for a given time period. For those with an education higher than  high school, we see that the data points are lighter for each hour, indicating that more females are having activities than males for a given time period.

### Question 3
```{r Q3_A}
janurary_20 = read_csv(file = "./data/Jan 2020 Citi.csv") %>%  #import data
  janitor::clean_names() %>% #clean column name
  mutate(
    time_period = "2020-01",
    month = 1,
    year = 2020
  )

janurary_24 = read_csv(file = "./data/Jan 2024 Citi.csv") %>%  #import data
  janitor::clean_names() %>% #clean column name
  mutate(
    time_period = "2024-01",
    month = 1,
    year = 2024
  )

july_20 = read_csv(file = "./data/July 2020 Citi.csv") %>%  #import data
  janitor::clean_names()  %>% #clean column name
  mutate(
    time_period = "2020-07",
    month = 7,
    year = 2020
  )

july_24 = read_csv(file = "./data/July 2024 Citi.csv") %>%  #import data
  janitor::clean_names()  %>% #clean column name
  mutate(
    time_period = "2024-07",
    month = 7,
    year = 2024
  )

citi = bind_rows(janurary_20, july_20, janurary_24, july_24)

citi %>% # type of rideable_type data in the data set
  count(rideable_type) %>%
  arrange(desc(n))

citi %>% # type of duration data in the data set --> trying to determine if time is in hr or min
  count(duration) %>%
  arrange(desc(duration))

citi %>% # type of member_casual data in the data set
  count(member_casual) %>%
  arrange(desc(n))
```

In 01/2020, there were a total of `r nrow(janurary_20)` records recorded. In 01/2024, there were a total of `r nrow(janurary_24)` records recorded. In 07/2020, there were a total of `r nrow(july_20)` records recorded. In 07/2024, there were a total of `r nrow(july_24)` records recorded. 

All four datasets have `r ncol(janurary_20)`. The dataset describes the type of ride an individual uses (classic or electric bike), which date of the week the ride was used, the total duration the ride was in minutes as there were values greater than 4, start and end station location, whether or not an individual was a casual or member rider.

```{r Q3_B}
# table showing the total number of rides in each combination of year and month separating casual riders and Citi Bike members
citi %>% 
  group_by(time_period, member_casual) %>% 
  summarize(total_count = n()) %>% 
  pivot_wider(
    names_from = "member_casual",
    values_from = total_count
  ) %>% 
  knitr::kable(digits = 1)
```

From the table above, it can be seen that as time progresses, both the number of casual riders and Citi Bike members increased. From 01/2020 to 07/2024, the number of casual riders have increased more than 10 folds and the number of Citi Bike members have increased more than 2 folds.

```{r Q3_C}
# table showing the 5 most popular starting stations for July 2024
july_24 %>% 
  group_by(start_station_name) %>% 
  summarize(total_count = n()) %>% 
  arrange(desc(total_count)) %>% 
  slice_head(n = 5) %>% # because the stations are on the columns, head() couldnt be used
  pivot_wider(
    names_from = "start_station_name",
    values_from = total_count
  )
```

The top 5 most population starting stations in 07/2024, was the following:

Pier 61 at Chelsea Piers
University Pl & E 14 St
W 21 St & 6 Ave
West St & Chambers St
W 31 St & 7 Ave

The 5 stations has a range of 146-163 riders in 07/2024.

```{r Q3_D}
# effects of day of the week, month, and year on median ride duration
citi %>% 
  mutate(
    weekdays = recode(weekdays,
      "Monday" = "1",
      "Tuesday" = "2",
      "Wednesday" = "3",
      "Thursday" = "4",
      "Friday" = "5",
      "Saturday" = "6",
      "Sunday" = "7"
      )
  ) %>% 
  group_by(weekdays, month, year) %>% 
  summarize(median_count = median(duration)) %>% 
  ggplot(aes(x = as.factor(weekdays), y = median_count, color = month)) +
  geom_point() + 
  facet_grid(. ~ as.factor(year)) +
  scale_x_discrete(labels = c("1" = "Monday", "2" = "Tuesday", "3" = "Wednesday", 
                               "4" = "Thursday", "5" = "Friday", "6" = "Saturday", 
                               "7" = "Sunday")) +
  labs(title = "Median Ride Duration By Day of Week",
       x = "Day of Week",
       y = "Median Ride Duration"
  )
```

There are more riders in July compared to January for both 2020 and 2024, and across all days of the week

```{r Q3_E}
citi_2024 = bind_rows(janurary_24, july_24)

citi_2024 %>%
  group_by(month, member_casual, rideable_type) %>% 
  ggplot(aes(x = duration, fill = interaction(member_casual, rideable_type))) +
  geom_histogram(position = "dodge") +
  facet_grid(. ~ month) +
  labs(title = "Rides Duration in Janurary_24 and July 2024",
       x = "Month",
       y = "Ride Duration"
  )
```

There is a higher duration of rides in July compared to January, which can be seen by comparing the two graphs. Both months seems to have a similar trend, where there are more electric bike riders than classic bike riders, as well as more members than casual individuals, especially in July. In both months, the peak of ride duration was around 15 miles, with relatively few individuals having a ride duration of over 50 miles.